\chapter{Selection I: Airplanes in War and Peace}

This homework exercise is about airplanes in war and peace, but it is
exactly the same as a model about how malaria affects the evolution of
sickle cell anemia or how typhoid fever affects the evolution of cystic
fibrosis. 

\section{Your Problem}

You have a fleet of two engine airplanes that just fly around in
peacetime and that engage the enemy during wartime.  A certain fraction
of engines that you receive are faulty, but you cannot check them before 
putting them on your airplanes.

The dynamics of your fleet in peacetime are as follows: all planes with two 
faulty engines are lost to you, ten percent of planes with one faulty
engine are lost to you, and five percent of planes with two good engines 
are lost to you each year.  Thus, if there are no faulty engines, your
fleet shrinks by five percent per year because of accidental losses.  We 
say that the \emph{mean fitness} of your airplanes is $0.95$ because you 
will have 95 planes at the end of each year for every 100 at the start
of the year.


\textbf{Problem 1. }What is the mean fitness of your fleet as a function of the
    fraction $q$ of engines that are faulty. Sketch a plot of mean
    fitness $\bar{w}$ versus $q$.

Say, for example, that ten percent of the engines you have are faulty.
Then, assuming that engines are put on planes at random, you expect that 
$0.10 \times 0.10 = 0.01$ or one percent of your planes will end up with 
two faulty engines and be lost, $2 \times 0.10 \times 0.90 = 0.18$ of 
your planes will have one faulty engine and suffer a ten percent loss
rate, while $0.9 \times 0.9 = 0.81$ of your planes will have two good
engines and suffer the five percent loss rate.  The mean fitness is then
$0 \times 0.01 + 0.90 \times 0.18 + 0.95 \times 0.81$.  You should
conclude from the exercise that faulty engines are bad for your fleet
unconditionally and that the more faulty engines the more airplanes you
will lose per year.  In genetics faulty engines would be called
``deleterious''. 

\textbf{Problem 2.} In peacetime what is the mean fitness of faulty
and good engines?  Overall mean fitness tells you how many planes you
have left at the end of the year.  Do the same thing for each kind of
engine. If this were genetics these would be called \emph{genic fitnesses.}
How do the engine fitnesses depend on the frequency of faulty engines?

\section{War}

A war occurs and the planes of your fleet are subject to a new source of 
loss: heat--seeking missiles.  A peculiar feature of these missiles is
that then cannot seek a plane with one faulty engine: instead they only
seek planes with two functioning engines.  The new parameters for your
airplane fleet in wartime are that all planes with two faulty engines
are lost, ten percent of planes with one faulty engine are lost (the
same as the rate during peace), and twenty percent of planes with two
good engines are lost.  This high rate of loss of planes with two good
engines is a consequence of enemy missiles.

Repeat all the above with the new wartime parameters.  Show that the
best fleet is one with ten percent of its engines faulty under this model in 
which engines are bolted on planes at random.  Sketch mean fitness and
the fitness of each type of engine as a function of the frequency of
faulty engines.

\section{Bonus problem: Wright versus Fisher} 

Do this problem if you are familiar with any simple computer
scripting language: basic, perl, python, awk, etc.


There is a new advantageous gene in a population of 5,000 breeding
organisms.  It is codominant, so that equation 3.2 of Gillespie
becomes

\begin{equation}
\delta p = \frac{pqs}{2\bar{w}}
\end{equation}

where

\begin{equation}
\bar{w} = 1-qs
\end{equation}

Assume the selective advantage $s$ is five percent, $s=0.05$.  What
will the frequency of the gene be in the total population of 10,000
after 100 generations under the two following scenarios:

\begin{itemize}

\item The population is panmictic so that the initial frequency of
the new mutant is just $1/10000$

\item The population is divided into ten demes and gene frequency
change occurs within a deme with negligible gene flow among them.  In
this case the initial frequency of the mutant is $1/1000$ in the deme
in which it occurs and zero elsewhere.

\end{itemize}
Under which scenario are there more copies of the new advantageous
mutant after 100 generations?





%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
