\chapter{Probability}
This lab is a set of exercises on probability.  For today only, you
will not need a computer: just paper, pencil, and maybe a pocket
calculator.  (Later labs will be different.)
\begin{enumerate}
\item You toss a fair coin twice.  What is the probability of
  observing two heads?

\item With the same coin experiment, what is the probability of a head
  and a tail (in that order)?

\item What is the probability of a tail and then a head?

\item What about a head and a tail in either order?

\item You toss a coin three times and observe 2 heads and a tail in
  some order.  In how many ways can this happen?  (In other words, how
  many sequences like ``HHT'' contains 2 ``H''s and one ``T''?)

\item What is the probability of observing two heads and a tail (order
  unspecified) in 3 tosses.

\item You toss two fair dice, one red and one black.  What is the
  probability that you observe either a red 4 or a black 6 (or both)?

\item What is the probability that you observe both a red 4 
  and a black 6?

\item You toss a fair coin 3 times.  You receive \$1 for each head and
  nothing for tails.  Let $X$ represent the number of dollars you
  receive.  What is (a)~the probability distribution of $X$, (b)~its
  mean, and (c)~its variance?

\item In JEPr, you saw several tables of counts.  In this exercise,
  you will construct one of your own by filling in the missing
  portions of the one below.
\begin{center}
\begin{tabular}{lccc}
       & \multicolumn{2}{c}{My mood}\\ \cline{2-3}
Weather& Happy & Sad& Sum\\ 
\hline 
Rain   &       &    & 100\\
Sun    &       &    & 100\\
\hline
Sum    &       &    & 200\\
\end{tabular}
\end{center}
So far, this table tells us that it rained on 100 days and was sunny
on another 100.  First fill in the blanks in the table, assuming that
I am more likely to be happy when it is sunny than when it is rainy.
Make sure that the rows for ``rain'' and ``sun'' each sum to 100.
Then use the bottom row to fill in column sums.  Now use your table to
calculate the \emph{unconditional} relative frequencies of (a)~sad
and (b)~rainy, then the \emph{conditional} relative frequencies of
(c)~sad given rainy and (d)~rainy given sad.  Is it true that
$f(\hbox{rain}|\hbox{sad}) = f(\hbox{rain})$?  If not, does this imply
that my mood influences the weather?

\item Consider a locus with two alleles, $A$ and $a$, whose
  frequencies are $p$ and $q=1-p$.  Suppose that you draw one gene
  copy from this population at random.  Let $X=1$ if that gene is a
  copy of $A$ and let $X=0$ otherwise.  What are (a)~the mean and
  (b)~the variance of $X$, as a function of $p$?

\item Consider a particular nucleotide site in the DNA of some
  hypothetical population.  In this population, 80\% of chromosomes
  have the nucleotide adenine (A) while 20\% have guanine (G).  In
  other words, the relative frequency of allele A is 0.80.  We take
  samples from this population, each of which consists of two
  chromosomes drawn independently and at random.  In each sample, we
  observe either AA, AG, or GG.  The number of A's in this sample is a
  random variable, which I will call $X$.  This variable can take
  values 0, 1, or 2.  What is the distribution of $X$?  What is the
  mean?  What is the variance?  The standard deviation?  Sketch a
  histogram of the distribution of $X$.

\item JEPr describes Bortkiewicz's data on the frequency of deaths
  caused by mule kicks in the Prussian army.  As you will recall,
  those data are approximately Poisson, and the mean number of deaths
  per corps-year was $\lambda = 0.61$.  What is the probability that
  there will be zero such deaths during a given year for a given
  corps?

\item Humans and fish differ in 68 amino acids out of 141 in alpha
  hemoglobin.  When they differ, there may have been undetected
  intermediate changes as they diverged during evolution.  If changes
  occur at random, what is the mean number of changes per amino acid
  site?  The ancestral lines of fish and humans diverged about 400
  million years ago.  Assuming that there have been 800 million years
  of independent evolution of these lines, what is the rate of amino
  acid substitution per year.  (This question is from a book by James
  Crow.  In answering it, assume that the number of changes separating
  human and fish is a Poisson random variable.  Use the data to estimate
  the mean $(\lambda)$ of this distribution.  To get started, ask
  yourself (a)~what is the probability that a Poisson random variable
  is equal to 0, and (b)~what fraction of amino acid sites are
  unchanged.) 

%\item Congenital adrenal hypoplasia (CAH) is a genetic disease that can
%  distort the external genetalia of females and can produce adrenal crises in
%  victims of either sex.  There is a test for the disease, but it is not
%  perfect.  5\% of affected individuals get negative test results, and 1\% of
%  unaffected individuals get positive results.  Among the Yupik eskimo, 1
%  individual in every 400 individuals has the disease.  
%\begin{enumerate}
%\item What is the probability that a Yupik eskimo \emph{has} the disease given
%  that   he got a \emph{positive} test result?
%\item What is the probability that a Yupik eskimo \emph{doesn't have} the
%  disease given that he got a \emph{negative} test result?
%\end{enumerate}
\end{enumerate}
